Do researchers preferentially collaborate with same-gendered colleagues?

### 2018-05-01 manuscript/Manuscript.Rmd
writing

# Methods
## Obtaining author lists and assigning gender
We used the dataset collected from PubMed by Holman et al. [-@Holman_2018]. Briefly, that dataset was created by downloading every single article indexed on PubMed, from the oldest articles up to the present (which was August 2016), and attempting to assign a gender to each author based on their given name, using the web service genderize.io. Each journal was assigned to one of 107 scientific disciplines (e.g. “Nephrology” or “Cell Biology”), using PubMed’s journal categorisations where available. Because the present study focuses on co-authorship, all single-author papers were removed from the dataset. To simplify the statistical analysis, we also discarded all papers for which we could not determine the gender of every author with at least 95% certainty. To mitigate Wahlund effects caused by variation in the gender ratio of researchers over time (see below), we also discarded all papers except those that were published 0-1 or 10-11 years before the PubMed data were collected (i.e. 20 August 2016). Lastly, we excluded journals with fewer than 50 suitable papers. This left us with a dataset from August 2015 - August 2016 containing `r sample.size$n2015[3]` papers with `r sample.size$n2015[4]` authorships, and a dataset from August 2005 - August 2006 containing `r sample.size$n2005[3]` papers and `r sample.size$n2005[4]` authorships. These papers came from `r sample.size$n2015[2]` and `r sample.size$n2005[2]` journals respectively, which were grouped into `r sample.size$n2015[1]` and `r sample.size$n2005[1]` research disciplines. There was a median of `r sample.size$n2015[5]` (`r sample.size$n2005[5]`) papers per journal, `r sample.size$n2015[6]` (`r sample.size$n2005[6]`) authors per journal, and `r sample.size$n2015[7]` (`r sample.size$n2005[7]`) authors per paper (the first number is for 2015-6, and the bracketed number for 2005-6).

@@

To address this, we also analysed every combination of journal and author country of affiliation for which we had enough data (i.e. 50 or more papers published in 2015-16). For simplicity, we restricted the dataset to only include papers for which Holman et al. [-@Holman_2018] had successfully identified the country of affiliation for all authors on the paper, and all authors shared the same country of affiliation. Restricting the dataset in this fashion produced enough data to measure $\alpha'$ for 325 combinations of journal and country (median: 70 papers and 273 authors per combination).

@@

$\alpha'$ was significantly higher in the sample from 2015-16 relative to 2005-6, but the difference in means was small (Figure S3; Effect of the fixed factor ‘time period’ in a linear mixed model of the data for all author positions: Cohen’s $d$ = 0.09 ± 0.04, $t_{948}$ = 2.51, p = 0.012). This result illustrates that gender homophily was slightly stronger in 2015-6 relative to 2005-6.

When comparing pairs of $\alpha'$ values estimated for the first and last authors for the same journals, we found that $\alpha'$ tended to be higher for first authors than for last authors (Figure S4; Effect of the fixed factor ‘Authorship position’ in a linear mixed model: Cohen’s d = 0.07 ± 0.02, $t_{1988}$ = 4.48, p < 0.0001). This suggests that the gender of the first author was a slightly stronger predictor of the remaining authors’ genders than the gender of the last author, i.e. the opposite of our prediction based on the hypothesis that senior scientists are causally responsible for homophily.

## Possible differences in the strength of homophily between disciplines
\autoref{alpha_histograms} illustrates the variance in journal homophily values ($\alpha'$) across scientific disciplines. All disciplines had a positive average $\alpha'$, although homophily appeared somewhat stronger in some disciplines than others (e.g. mean $\alpha'$ was 0.12 ± 0.02 for Urology journals, and 0.03 ± 0.01 for Veterinary Medicine journals; \autoref{alpha_histograms}, Table S3). However, there was little evidence for strong differences in $\alpha'$ between disciplines: the random factor ‘discipline’ explained at most 1% of the variance in $\alpha'$ in the two linear mixed models described in the previous section (see \autoref{alpha_histograms} and mixed models in Online Supplementary Material). This implies that the processes causing positive $\alpha'$ are similarly strong in all the disciplines we examined.

Reassuringly, we found no support for our prediction that journals publishing on a wide range of topics should have higher $\alpha'$ values than more specialised journals, due to the Wahlund effect. Specifically, the journal category “Multidisciplinary” - which includes journals like _PLoS ONE_, _Nature_, _Science_, and _PNAS_ - did not have notably elevated $\alpha'$ (\autoref{alpha_histograms}). This result suggests that our estimates of homophily are not notably inflated by the presence of disparate research topics (with variable author gender ratios) being published within individual journals.

## Relationship between gender homophily and gender ratio
One might predict that researchers would be more likely to seek out same-gendered colleagues in strongly gender-biased disciplines (e.g. Surgery or Nursing), relative to disciplines with a comparatively gender-balanced workforce (e.g. Psychiatry). However, we instead found evidence of the reverse: there was a positive, non-linear relationship between the overall gender ratio of all authors publishing in a particular journal [as estimated by @Holman_2018], and the estimated value of $\alpha'$ for all authors and for first authors (\autoref{alpha_gender_ratio}). That is, journals with a balanced or female-biased author gender ratio tended to have higher $\alpha'$ than journals with a male-biased author gender ratio (GAM smooth terms p < 0.001, see Online Supplementary Material). The relationship was not statistically significant when $\alpha'$ was calculated for last authors (GAM, p = 0.142), though the trend appeared similar (\autoref{alpha_gender_ratio}).

@@

## Analysis correcting for differences between countries
When we restricted the analysis to only include authors with affiliations from a single country, we found statistically significant homophily for 72 of the 325 journal-country combinations tested (64 unique journals and 18 unique countries), and no significant evidence of heterophily (Figures S5-S6). Additionally, the values of $\alpha'$ calculated for each journal-country combination were mostly very similar to the $\alpha'$ values calculated for the journal as a whole (i.e. when pooling papers from different countries); the average difference in $\alpha'$ was 0.002 (Figure S7). These results suggest that our main analysis’ findings of widespread homophily are not driven by a Wahlund effect resulting from gender differences between countries.

# Discussion
We found evidence that researchers tend to publish with same-gendered coauthors, even after implementing stringent controls for ‘spurious’ homophily resulting from Wahlund effects (\autoref{wahlund_plot}). Our data therefore reaffirm earlier conclusions (Ferber and Teiman 1980; McDowell and Smith 1992; Crow and Smykla 2015; Ghiasi et al. 2015; Zettler et al. 2016; Fahmy and Young 2016; Jadihi et al. 2017; Araujo and Fontainha 2017a; Teele and Thelen 2017) that research collaborations occur between same-gendered colleagues substantially more often than expected if collaborators are selected without regard to their gender. Many journals had $\alpha'$ values over 0.1, meaning that the gender ratio of the average man's coauthors is skewed towards men by >10%, relative to the gender ratio of a woman's coauthors (or equivalently, that the gender ratio of women's coauthors is >10% more female than that of men's). Few journals had $\alpha'$ values below zero, and fewer still had significantly negative $\alpha'$ values, suggesting that researchers rarely seek out collaborators of the opposite gender (or at least, that they do so less often than they seek out same-gendered colleagues).

We found no evidence that gender homophily is restricted to particular disciplines: instead, there was evidence of homophily across many diverse research disciplines. We also found that gender homophily was marginally stronger in 2015-2016, relative to 2005-2006. This result suggests that recent increases in the number of women working in STEMM, as well as recent efforts to recruit, retain, and celebrate women in STEMM, have (if anything) strengthened gender homophily rather than weakening it.

>>>>

# Methods
## Obtaining author lists and assigning gender
We used the dataset of PubMed author lists from Holman et al. [-@Holman_2018]. Briefly, that dataset was created by downloading every single article indexed on PubMed, from the oldest articles up to the present (which was August 2016), and attempting to assign a gender to each author based on their given name, using the web service genderize.io. Each journal was assigned to one of 107 scientific disciplines (e.g. “Nephrology” or “Cell Biology”), using PubMed’s journal categorisations where available. Because the present study focuses on co-authorship, all single-author papers were removed from the dataset. To simplify the statistical analysis, we also discarded all papers for which we could not determine the gender of every author with at least 95% certainty. To mitigate Wahlund effects caused by variation in the gender ratio of researchers over time (see below), we also discarded all papers except those that were published 0-1 or 10-11 years before the PubMed data were collected (i.e. 20 August 2016). Lastly, we excluded journals with fewer than 50 suitable papers. This left us with a dataset from August 2015 - August 2016 containing `r sample.size$n2015[3]` papers with `r sample.size$n2015[4]` authorships, and a dataset from August 2005 - August 2006 containing `r sample.size$n2005[3]` papers and `r sample.size$n2005[4]` authorships. These papers came from `r sample.size$n2015[2]` and `r sample.size$n2005[2]` journals respectively, which were grouped into `r sample.size$n2015[1]` and `r sample.size$n2005[1]` research disciplines. There was a median of `r sample.size$n2015[5]` (`r sample.size$n2005[5]`) papers per journal, `r sample.size$n2015[6]` (`r sample.size$n2005[6]`) authors per journal, and `r sample.size$n2015[7]` (`r sample.size$n2005[7]`) authors per paper (the first number is for 2015-6, and the bracketed number for 2005-6).

@@

To address this, we also analysed every combination of journal and author country of affiliation for which we had enough data (i.e. 50 or more papers published in 2015-16). For simplicity, we restricted the dataset to only include papers for which Holman et al. [-@Holman_2018] had identified the country of affiliation for all authors on the paper, and all authors shared the same country of affiliation. Restricting the dataset in this fashion produced enough data to measure $\alpha'$ for 325 combinations of journal and country (median: 70 papers and 273 authors per combination).

@@

$\alpha'$ was significantly higher in the sample from 2015-16 relative to 2005-6, but the difference in means was small (Figure S3; Effect of the fixed factor ‘time period’ in a linear mixed model of the data for all author positions: Cohen’s $d$ = 0.09±0.04, $t_{948}$ = 2.51, p = 0.012).

When comparing pairs of $\alpha'$ values estimated for the first and last authors for the same journals, we found that $\alpha'$ tended to be higher for first authors than for last authors (Figure S4; Effect of the fixed factor ‘Authorship position’ in a linear mixed model: Cohen’s $d$ = 0.07±0.02, $t_{1988}$ = 4.48, p < 0.0001). This suggests that the gender of the first author was a slightly stronger predictor of the remaining authors’ genders than the gender of the last author, i.e. the opposite of our prediction based on the hypothesis that senior scientists are causally responsible for homophily.

## Possible differences in the strength of homophily between disciplines
\autoref{alpha_histograms} illustrates the variance in journal homophily values ($\alpha'$) across scientific disciplines. All disciplines had a positive average $\alpha'$, although homophily appeared somewhat stronger in some disciplines than others (e.g. mean $\alpha'$ was 0.12±0.02 for Urology journals, and 0.03±0.01 for Veterinary Medicine journals; \autoref{alpha_histograms}, Table S3). However, there was little evidence for strong differences in $\alpha'$ between disciplines: the random factor ‘discipline’ explained at most 1% of the variance in $\alpha'$ in the two linear mixed models described in the previous section (see \autoref{alpha_histograms} and mixed models in Online Supplementary Material). This implies that the processes causing positive $\alpha'$ are similarly strong in all the disciplines we examined.

There was no indication that journals publishing on a wide range of topics have higher $\alpha'$ values than more specialised journals, due to the Wahlund effect. For example, the journal category “Multidisciplinary” - which includes journals like _PLoS ONE_, _Nature_, _Science_, and _PNAS_ - did not have notably elevated $\alpha'$ (\autoref{alpha_histograms}). This result suggests that our estimates of homophily, and estimates from earlier studies, are not notably inflated by the presence of disparate research topics (with variable author gender ratios) being published within individual journals.

## Relationship between gender homophily and gender ratio
We next tested whether researchers are more or less likely to seek out same-gendered colleagues in strongly gender-biased disciplines (e.g. Surgery or Nursing), relative to disciplines with a comparatively gender-balanced workforce (e.g. Psychiatry). We found a positive, non-linear relationship between the overall gender ratio of all authors publishing in a particular journal [as estimated by @Holman_2018], and the estimated value of $\alpha'$ for all authors and for first authors (\autoref{alpha_gender_ratio}). Journals with a balanced or female-biased author gender ratio tended to have higher $\alpha'$ than journals with a male-biased author gender ratio (GAM smooth terms p < 0.001, see Online Supplementary Material). The relationship was not statistically significant when $\alpha'$ was calculated for last authors (GAM, p = 0.142), though the trend appeared similar (\autoref{alpha_gender_ratio}).

@@

## Analysis correcting for differences between countries
When we restricted the analysis to only include authors with affiliations from a single country, we found statistically significant homophily for 72 of the 325 journal-country combinations tested (64 unique journals and 18 unique countries), and no significant evidence of heterophily (Figures S5-S6). Additionally, the values of $\alpha'$ calculated for each journal-country combination were mostly very similar to the $\alpha'$ values calculated for the journal as a whole (i.e. when pooling papers from different countries); the average difference in $\alpha'$ was 0.002 (Figure S7). These results suggest that our findings of widespread homophily in the main analysis cannot solely be driven by a Wahlund effect resulting from gender differences between countries.

# Discussion
We found evidence that researchers tend to publish with same-gendered coauthors, despite implementing stringent controls for ‘spurious’ homophily resulting from Wahlund effects (\autoref{wahlund_plot}). Our data therefore reaffirm earlier conclusions (Ferber and Teiman 1980; McDowell and Smith 1992; Crow and Smykla 2015; Ghiasi et al. 2015; Zettler et al. 2016; Fahmy and Young 2016; Jadihi et al. 2017; Araujo and Fontainha 2017a; Teele and Thelen 2017) that research collaborations occur between same-gendered colleagues substantially more often than expected if collaborators are selected without regard to their gender. Many journals had $\alpha'$ values over 0.1, meaning that the gender ratio of the average man's coauthors is skewed towards men by >10%, relative to the gender ratio of a woman's coauthors. Few journals had $\alpha'$ values below zero, and fewer still had significantly negative $\alpha'$ values, suggesting that it is comparatively rare for researchers to preferentially select opposite-gendered collaborators.

We found no evidence that gender homophily is restricted to particular disciplines: instead, there was evidence of homophily across many diverse research disciplines. We also found that gender homophily was marginally stronger in 2015-2016, relative to 2005-2006. This result suggests that recent increases in the number of women working in STEMM, as well as recent efforts to recruit, retain, and celebrate women in STEMM, have co-occurred with rise in the frequency of same-gender collaborations, not a fall. We speculate that this trend might result, at least in part, from the increasing number of women working in senior positions in STEMM over the last 10 years (DATA?). Because many co-authorships involve a junior and a senior researcher (e.g. a student and professor), and the gender ratio is typically more skewed among senior staff (Shaw and Stanton ref), there will be more opposite-gender collaborations than expected if the gender ratio were the same at both career stages. Thus, the observed increase in homophily might be due to the greater numbers of women working in leading roles.
